---
title: 'Excourse: Variability of signature activities'
author: "Drews RM et al."
date: "20/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(reshape2)
library(ggplot2)
library(ggthemes)
library(patchwork)
library(ComplexHeatmap)
theme_set(theme_tufte(base_size = 12, base_family = "Arial"))

BASE="~/cnsigs2_revisions"
ORI=file.path(BASE, "CINsignatures/Signature_Compendium_v5_Cosine-0.74_Activities.rds")
ASCAT25=file.path(BASE, "out/4_TCGA_PCAWG_Exposures_to_TCGA_Signatures_ASCAT25.rds")
ASCAT200=file.path(BASE, "out/4_TCGA_PCAWG_Exposures_to_TCGA_Signatures_ASCAT200.rds")

mOri = readRDS(ORI)
m25 = readRDS(ASCAT25)
m200 = readRDS(ASCAT200)


```

## Plan

So far derived signature activities for ASCAT with penalties 25, 70 and 200. Let's see how close they are.

```{r test}

mOri = mOri[ rownames(mOri) %in% rownames(m25) & rownames(mOri) %in% rownames(m200), ]
m25 = m25[ rownames(m25) %in% rownames(mOri) & rownames(m25) %in% rownames(m200), ]
m200 = m200[ rownames(m200) %in% rownames(m25) & rownames(m200) %in% rownames(mOri), ]

identical(rownames(mOri), rownames(m25))
identical(rownames(m200), rownames(m25))

# Simple test - subtract values from each other
Heatmap(mOri - m25, cluster_columns = FALSE, show_row_names = FALSE, name = "70 - 25")
summary(mOri - m25)

Heatmap(mOri - m200, cluster_columns = FALSE, show_row_names = FALSE, name = "70 - 200")
summary(mOri - m200)

```

The signature activities seem quite different from each other. What if we look only into one signature? How many samples change to zero and the other way around?


```{r zeros in CS16}

## Samples with zero sig activity under ASCAT 70
zeroIDs = rownames(mOri)[ mOri[,16] == 0 ]
length(zeroIDs)

## Distribution of values for CS16 under ASCAT 25 for the same samples
vals25 = m25[ rownames(m25) %in% zeroIDs, 16 ]
sum(vals25 == 0)
summary(vals25)

## Distribution of values for CS16 under ASCAT 200 for the same samples
vals200 = m200[ rownames(m200) %in% zeroIDs, 16 ]
sum(vals200 == 0)
summary(vals200)

```

For ASCAT25, roughly 5% of samples with previously zero activity switched to a non-zero activity now. For ASCAT200, around 20% of zero samples switched to a non-zero state. 

Let's test all signatures.

```{r zero changes in all sigs}

allSigs = colnames(mOri)
lZeros = lapply(allSigs, function(thisSig) {
  
  ## Samples with zero sig activity under ASCAT 70
  zeroIDs = rownames(mOri)[ mOri[,thisSig] == 0 ]
  numZero = length(zeroIDs)
  
  ## Distribution of values for CS16 under ASCAT 25 for the same samples
  vals25 = m25[ rownames(m25) %in% zeroIDs, thisSig ]
  numZero25 = sum(vals25 == 0)
  
  ## Distribution of values for CS16 under ASCAT 200 for the same samples
  vals200 = m200[ rownames(m200) %in% zeroIDs, thisSig ]
  numZero200 = sum(vals200 == 0)

  out = c(thisSig, numZero, numZero25, numZero200)
  return(out)
  
})

dtZero = data.table(do.call(rbind, lZeros))
colnames(dtZero) = c("Sig", "Zeros70", "OfThoseZero25", "OfThoseZero200")

dtZeroMelt = melt(dtZero, id.vars = "Sig")
dtZeroMelt$value = as.numeric(dtZeroMelt$value)
dtZeroMelt$Sig = factor(as.character(dtZeroMelt$Sig), levels = paste0("CS", 1:17))

ggplot(dtZeroMelt, aes(x = Sig, y = value, fill = variable)) + 
  geom_bar(stat = "identity", position = "dodge")

```

What next? ASCAT 50 and 100 for all of TCGA and hope that the results are closer? Estimate probability of being non-zero when zero? Take values of samples that switch from zero to non-zero as non-interpretable signature activities (could be quite high)? Monte carlo simulations on noise?



